Binning Strategies and Related Loss for Large Data

### 2015-05-02 Paper.Rnw
Lots of edits.  Most importantly an emphasis that we choose to bin frequencies to make frequency loss explicit. We experience this loss implicitly because of our inability to reverse the mapping of frequency to continuous shade through perception.

In this paper we describe first the problem scatterplots are exposed to in large-data situations. We discuss different binning algorithms use the construct binned scatterplots and the  {\it loss of information} inherent to binning. We will then explore the binning properties through simulation and real-data case studies. We conclude with several practical suggestions for binning specifications for creating binned scatterplots that have desirable visual properties.

@@

Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.  A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, \hh{such as color or size (or 3D renderings, which we do not want to recommend at this point)} to provide joint density information within each grid cell, known as a tile.  A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) at the bin center, similar to a two-dimensional histogram viewed from above. A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.  Bubble plots were first used by William Playfair \citep{playfair, playfair2}.

@@

Binning algorithms used in making distributional approximations can be traced back to Pearson's work with the binomial approximation to the normal, where he mentions the need to define an origin and binwidth for segmenting the normal distribution \citep{Pearson1895}. More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations (\citealp{scott1979} and \citealp{scott1992}). \hh{XXX the multivariate book is going beyond one dimension}

@@

\ktm{
\begin{itemize}
\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a set of tiles centered at $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$. By displaying frequency information using tiles that are visually centered at the bin centers causes a loss of information about the exact location of the points. \hh{XXX the frequency information is not what causes the loss}

\item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not rendered precisely. A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell)$, for spatial bins $\ell \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies. This is analogous to the fundamental problem for over-plotted standard scatterplots where frequency is masked by other points. 
\hh{XXX ok, I don't quite agree with the description of frequency loss here. (a) the rendering on a continuous color scheme IS precise. but on a cognitive level we are not able to read the rendered values at the same precision. This effectively intorduces a second stage of binning ... then we are on the same page again.}
\end{itemize}
}

\hh{XXX what are you referring to by `it' in the next sentence?}
Note that even though these losses may turn out to be substantial, it is, in fact, a huge gain with respect to the original scatterplot, where information is masked in large data situations due to over-plotting of points. The idea of loss from one-dimensional binning was explored by Scott using mean squared error as the loss function to be optimized by the choice of the number of bins in  the construction of histograms \citep{scott1979}. He later extended this discussion to two-dimensional binning, where he compared the mean squared error loss for hexagonal, rectangular and triangular binning; finding that hexagonal and rectangular binning performed similarly, both far superior to triangular binning \citep{scott1992}. We will take a similar approach to quantifying loss resulting from the binning algorithms detailed in Section~\ref{GenBinning} above.

@@

\ktm{The net spatial loss is a euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded. \hh{XXX isn't this true for spatial loss in general and not just the net spatial loss? }  If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.  However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard deviations. }

@@

If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency. \ktm{As bin size becomes increasingly fine, the display will place the visual emphasis for the frequency information on more precise spatial locations but \hh{XXX what do you mean by bin density?} bin density have higher sampling variability due to the smaller number of observation per bin.} Choosing bin size for a binned scatterplot carries a trade-off between unbiasedness and consistency of the visual density estimate. For efficiency of density estimation, it may be tempting to claim that a minimally binning, with only one unique coordinate pair per bin, is an efficient estimator because it minimizes spatial loss. Minimally binned scatterplots are \textit{spatially} efficient, but density estimation requires the combination of spatial and frequency information, each of which have separate loss functions. Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.

@@

\hh{XXX please re-phrase the paragraph such that it reflects that we do not WANT to bin colors, but that this effectively happens when we look at a figure, because of the limitations of human perception. We use the binning to reflect what happens in practice. The binning then allows us to measure what kind of loss occurs because of the limitations of the human cognitive system.}

\ktm{The reduced binned data from spatial binning contains the center and count information for all bins. The bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot. }
Figure~\ref{fig:StandardFreqBinning4710none} displays binned scatterplots with varying numbers of  standard binned frequency groups. If we attempt to discern differences between similarly shaded tiles: it is trivial when only four shades exist, it becomes more difficult -- but still possible -- at seven bins, and at ten frequency bins we are hardly able to discriminate better than in continuous shading. This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}. Our exploration of frequency loss will focus on frequency binning with at most ten bins.

@@

Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins. However, using both -- quantile log frequency binning -- is not a recommended approach \hh{XXX not recommended is too strong. It actually does not help to do both, as can be seen in the figure}. The frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles. 
In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.

@@

\ktm{The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot with a continuous shading of the raw bin frequencies. \hh{ XXX the legend needs fixing.  the numbers are running into one another. XXX be careful with `density' of the bins - the bins show frequencies not densities.} We quickly realize that there is a large number of pitchers that played very few games and had very few strikeouts, however this renders it nearly impossible to differentiate the density structure across the majority of bins which are filled with higher quality pitchers who played many games and earned many strikeouts. In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts. Quantile frequency binning using four shade groups (center plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) allows us to visualizes the quartiles of the bin densities. Alternatively, standard log frequency binning using four shade groups (rightmost plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) uses the log transformation to diminish the visual impact of high density bins then visualizes density on the log scale. Each frequency binning approach adds a layer of complexity to the interpretation of the shade, however they both effectively emphasize the two ridges in the density structure. }

>>>>

In this paper we describe first the problem scatterplots are exposed to in large-data situations. We discuss different binning algorithms use the construct binned scatterplots and the  {\it loss of information} inherent to binning. \ktm{We will then explore the effects of binning specification on the properties of binned scatterplots through simulation and real-data case studies.} We conclude with several practical suggestions for binning specifications for creating binned scatterplots that have desirable visual properties.

@@

Histograms are a simple example of a plot that can be built using binned aggregations of the data; in their case the bin locations and bin counts act as a set of sufficient statistics necessary to reconstruct the plot.  A natural extensions of histograms to higher dimensions is to form a tessellated grid on a two dimensional Cartesian plane using some other attribute, such as color or size (or 3D renderings, which we do not want to recommend at this point) to provide joint density information within each grid cell, known as a tile.  A {\it binned scatterplot} uses shading to provide frequency information, with tiles (rather than bars in a histogram) at the bin center, similar to a two-dimensional histogram viewed from above. A \textit{bubble plot} is a binned data plot that scales the size of a filled circle in proportion to frequency.  Bubble plots were first used by William Playfair \citep{playfair, playfair2}.

@@

Binning algorithms used in making distributional approximations can be traced back to Pearson's work with the binomial approximation to the normal, where he mentions the need to define an origin and binwidth for segmenting the normal distribution \citep{Pearson1895}. More recently Scott has presented discussion on the importance of binning specification in the creation of histograms to appropriately display one dimensional density approximations \citep{scott1979}. \citet{scott1992} extends to the properties of multivariate binning strategies.

@@

\begin{itemize}
\item {\it Spatial Loss}, $L^\text{S}$, occurs when points $(x_i, y_i)$ for observations $i \in \{1,\dots,n\}$ in the data set are reduced to a set of tiles centered at $(x_\ell^\ast, y_\ell^\ast)$ for bins $\ell \in \{1,\dots,\mathscr{L}\}$. By displaying frequency information using shaded tiles instead of individual points there is a loss of information about the exact location of the points. %\hh{XXX the frequency information is not what causes the loss}
\ktm{
\item {\it Frequency Loss}, $L^\text{F}$, occurs when bin counts, $c_\ell \in \{1,\dots,\mathscr{L}\}$ are not mapped to a continuous shading scale. While shade can be \textit{rendered} continuously in HSV color space, thus representing frequency exactly, this information is extracted via imperfect human perception. A discrete color scale may be used to aid the ability to read a binned scatterplot, however this is effectively a second stage of binning where shade displays binned frequencies, $b_C(c_\ell)$, for spatial bins $\ell \in \{1,\dots,\mathscr{L}\}$, thus losing precision in the display of the true frequencies.
}
\end{itemize}

\hh{XXX ok, I don't quite agree with the description of frequency loss here. (a) the rendering on a continuous color scheme IS precise. but on a cognitive level we are not able to read the rendered values at the same precision. This effectively intorduces a second stage of binning ... then we are on the same page again.} \ktm{Is this new description any better?}

Note that even though the losses from creating a binned scatterplot may turn out to be substantial there a huge gain with respect to the original scatterplot, where information can be masked in large data situations due to over-plotting of points. The idea of loss from one-dimensional binning was explored by Scott using mean squared error as the loss function to be optimized by the choice of the number of bins in  the construction of histograms \citep{scott1979}. He later extended this discussion to two-dimensional binning, where he compared the mean squared error loss for hexagonal, rectangular and triangular binning; finding that hexagonal and rectangular binning performed similarly, both far superior to triangular binning \citep{scott1992}. We will take a similar approach to quantifying loss resulting from the binning algorithms detailed in Section~\ref{GenBinning} above.

@@

\ktm{The spatial loss is a euclidean distance, but the units affiliated with this distance are based on the units on which each variable is recorded. If the two variables in the binned scatterplot share the same units this leads to direct interpretability of the spatial loss.  However, if the two variables do not share the same units or magnitude of values it is advisable to standardize the variables prior to binning, thus the spatial loss is more universally interpretable as a distance in units of standard deviations. }

@@

If we view the binned scatterplot as a visual estimator of the bivariate density then we may consider a few desirable properties of estimators: unbiasedness, consistency and efficiency. \ktm{Binned scatterplots are inherently bias displays of spatial information as they shift visual emphasis from true location of individual points to the geometric centers of tiles. However, as bin size becomes increasingly fine, the plot will place visual emphasis more precise spatial locations. When minimal binning occurs -- with one unique coordinate pair at the center of each bin -- the density estimate then spatially exact; thus making the binned scatterplot a consistent visual estimator. We may also consider minimally binning as a spatially efficient estimator because it minimizes spatial loss in the visual density estimator. It is worth noting that a density estimate requires the combination of spatial and frequency information and that the estimation properties were considered only through the lens of spatial loss. This makes the assumption that frequency information is rendered through a precise continuous mapping of bin frequencies to bin shades. %Based on the conflicting effects on the estimation properties, of bin size can be seen as a smoother for density estimate that can be adjusted to cut through the noise in the data, but as with all data smoothers caution should be used so that fine signal is not destroyed.  
}

@@

%\hh{XXX please re-phrase the paragraph such that it reflects that we do not WANT to bin colors, but that this effectively happens when we look at a figure, because of the limitations of human perception. We use the binning to reflect what happens in practice. The binning then allows us to measure what kind of loss occurs because of the limitations of the human cognitive system.}

\ktm{The reduced binned data from spatial binning contains the center and count information for all bins. The bin frequencies may be mapped continuously to a precisely rendered shade, however it is niave to believe that human perception will be able to perfectly extract that information. There is implicitly loss of frequency information occuring when the shade of a tile is visually mapped back to a frequency through the use of a shading scale index. Bin frequencies may themselves be binned in order to discretize the color scale for the binned scatterplot, thus making the loss explicit.

Figure~\ref{fig:StandardFreqBinning4710none} displays binned scatterplots with varying numbers of  standard binned frequency groups. If we attempt to discern differences between similarly shaded tiles: it is trivial when only four shades exist, it becomes more difficult -- but still possible -- at seven bins, and at ten frequency bins we are hardly able to discriminate better than in continuous shading. This aligns with Healey and Enn's theory on the number of discernible colors \citep{colorperception}. Our exploration of frequency loss will focus on frequency binning with at most ten bins because above this we experience implicit frequency from perceptual bounds that are not well reflected in the explicitly defined frequency loss. }

@@

Log transforming the frequencies prior to binning and using quantile based binning on the raw frequencies are two methods for dealing with the same problem for binned scatterplots: heavily right skewed bin counts where dense bins visually overshadow any structure in low density bins. However, using both -- quantile log frequency binning -- is not a recommended approach \hh{XXX not recommended is too strong. It actually does not help to do both, as can be seen in the figure}. \ktm{Frequency groups for quantile binning are invariant to the log transformation because a monotone transformation does not affect groupings based on quantiles, thus it is preferable leave the frequencies untransformed before quantile binning for better contextual interpretability.} In the common situation where bin frequencies are heavily skewed, the choice is between quantile frequency binning and standard log frequency binning.

@@

\ktm{The leftmost plot in Figure~\ref{fig:BaseballBinnedScatters2X10} displays the binned scatterplot with a continuous shading of the raw bin frequencies. We quickly realize that there is a large number of pitchers that played very few games and had very few strikeouts, however this renders it nearly impossible to differentiate the density structure across the majority of bins which are filled with higher quality pitchers who played many games and earned many strikeouts. In this case there are two frequency binning options that we can employ to deal with this skewness of the distribution of bin counts. Quantile frequency binning using four shade groups (center plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) allows us to visualizes the quartiles of the bin densities. Alternatively, standard log frequency binning using four shade groups (rightmost plot of Figure~\ref{fig:BaseballBinnedScatters2X10}) uses the log transformation to diminish the visual impact of high density bins then visualizes density on the log scale. Each frequency binning approach adds a layer of complexity to the interpretation of the shade, however they both effectively emphasize the two ridges in the density structure. }
